---
title: "High-Dimensional Classification with PCA"
author: 'Carolyne Chepkemboi '
output:
  html_document:
    df_print: paged
---

### Data Loading and Exploration

Load required libraries for this notebook.

```{r}
library(ggplot2)
library(dplyr)
# custom set select() (since MASS is loaded, overriding select())
select <- dplyr::select
library(FactoMineR)
library(factoextra)
library(class)
library(caret)
library(plotly)
library(knitr)
set.seed(6348)
```

Load the Human Activity Recognition Using Smartphones dataset [Human Activity Recognition Using Smartphones dataset](https://archive.ics.uci.edu/dataset/240/human+activity+recognition+using+smartphones) into memory.

The inner UCI HAR Dataset folder, as downloaded from the UC Irvine repository and unzipped, should be placed in the same directory as this file.

Alternative solution: unzip the data, drag the "UCI HAR Dataset" subfolder into documents, set wd as setwd("~/Documents"), and the code should compile.

```{r}
# Training data/labels
X_train_raw <- read.table("C:/Users/carolynechep/Desktop/human+activity+recognition+using+smartphones/UCI HAR Dataset/UCI HAR Dataset/train/X_train.txt")
y_train_raw <- read.table("C:/Users/carolynechep/Desktop/human+activity+recognition+using+smartphones/UCI HAR Dataset/UCI HAR Dataset/train/y_train.txt")

# Test data/labels
X_test_raw <- read.table("C:/Users/carolynechep/Desktop/human+activity+recognition+using+smartphones/UCI HAR Dataset/UCI HAR Dataset/test/X_test.txt")
y_test_raw <- read.table("C:/Users/carolynechep/Desktop/human+activity+recognition+using+smartphones/UCI HAR Dataset/UCI HAR Dataset/test/y_test.txt")
```

```{r}
dim(X_train_raw) %>% print
dim(X_test_raw) %>% print

features <- dim(X_train_raw)[2]
train_len <- dim(X_train_raw)[1]
test_len <- dim(X_test_raw)[1]

X_train_raw[1:5,1:3] %>% print
y_train_raw[1:30,] %>% print
```

Rename the columns of the features dataframes to the given description of the corresponding feature.

```{r}
feature_descs <- read.table("C:/Users/carolynechep/Desktop/human+activity+recognition+using+smartphones/UCI HAR Dataset/UCI HAR Dataset/features.txt")
colnames(feature_descs) <- c("index", "desc")
colnames(X_train_raw) <- feature_descs$desc
colnames(X_test_raw) <- feature_descs$desc

generic_colnames <- paste("V", sapply(1:features, toString), sep = "") 
```

```{r}
X_train_raw[1:5,1:3] %>% print
```

Add a column to the labels dataframes with the corresponding activity name for ease of interpretability.

```{r}
classes <- read.table("C:/Users/carolynechep/Desktop/human+activity+recognition+using+smartphones/UCI HAR Dataset/UCI HAR Dataset/activity_labels.txt")

y_train <- y_train_raw %>%
  mutate(index = V1,
    activity = classes$V2[index]) %>%
  select(index, activity)

y_test <- y_test_raw %>%
  mutate(index = V1,
    activity = classes$V2[index]) %>%
  select(index, activity)
```

```{r}
y_train %>% count(activity) %>% print
```

Scale and center the raw features. We can't use `scale()` since we need to scale the test data with respect to the means and standard deviations of the training data.

```{r}
# Calculate the mean and sd of each raw feature
X_train_raw_mean <- colMeans(X_train_raw)
X_train_raw_sd <- apply(X_train_raw, 2, sd)

# Subtract the mean from each feature, then divide by the sd
X_train <- X_train_raw %>% 
  sweep(2, X_train_raw_mean, "-") %>%
  sweep(2, X_train_raw_sd, "/")

X_test <- X_test_raw %>% 
  sweep(2, X_train_raw_mean, "-") %>%
  sweep(2, X_train_raw_sd, "/")
```

```{r}
X_train[1:5,1:3] %>% print
```
### Exploratory Data Analysis
### Activity Class Distribution
```{r}
library(ggplot2)

y_train %>%
  count(activity) %>%
  ggplot(aes(x = activity, y = n)) +
  geom_bar(stat = "identity", fill = "#2C7FB8", width = 0.7) +
  theme_minimal(base_size = 14) +
  labs(title = "Class Distribution (Training Set)",
       x = "Activity", y = "Count") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```
### Feature Density Distributions
```{r}
library(tidyr)
X_sample <- X_train_raw[, 1:6]

X_long <- X_sample %>%
  as.data.frame() %>%
  pivot_longer(cols = everything())

ggplot(X_long, aes(x = value)) +
  geom_density(fill = "#2C7FB8", alpha = 0.6) +
  facet_wrap(~name, scales = "free") +
  theme_minimal() +
  labs(title = "Feature Density Distributions")
```
### Correlation Heatmap
```{r}
library(corrplot)

cor_mat <- cor(X_train_raw[, 1:10])  # subset for readability

corrplot(cor_mat, method = "color", tl.cex = 0.5)
```
### Standardization
Scale and center the raw features of training data using sweep()
```{r}
# Calculate the mean and sd of each raw feature
X_train_raw_mean <- colMeans(X_train_raw)
X_train_raw_sd <- apply(X_train_raw, 2, sd)

# Subtract the mean from each feature, then divide by the sd
X_train <- X_train_raw %>% 
  sweep(2, X_train_raw_mean, "-") %>%
  sweep(2, X_train_raw_sd, "/")

X_test <- X_test_raw %>% 
  sweep(2, X_train_raw_mean, "-") %>%
  sweep(2, X_train_raw_sd, "/")
```

```{r}
X_train[1:5,1:3] %>% print
```

### Additional EDA

Check if the data has missing values to handle

```{r}
sum(is.na(X_train_raw))
sum(is.na(X_test_raw))
```

Summary statistics and histogram of standard deviations to corroborate scaling

```{r}
standard_deviations <- apply(X_train_raw, 2, sd)
summary(standard_deviations)
IQR(standard_deviations)
hist(standard_deviations,
     main = "Standard Deviations Distributions",
     xlab = "Standard Deviations of Features",
     col = "#2f4569")
```

Number and percentage of absolute and raw correlations greater than 0.9.

```{r}
correlations <- cor(X_train_raw)[upper.tri(cor(X_train_raw))]
sum(abs(correlations) > 0.9)
sum(correlations > 0.9)
sum(abs(correlations) > 0.9) / ((features * (features-1)) / 2)
sum(correlations > 0.9) / ((features * (features-1)) / 2)
```

Percentage of absolute and raw correlations greater than 0.8.

```{r}
sum(abs(correlations) > 0.8) / ((features * (features-1)) / 2)
sum(correlations > 0.8) / ((features * (features-1)) / 2)
```

Distribution of pairwise correlations.

```{r}
hist(correlations,
     main = "Distribution of Unique Pairwise Correlations",
     xlab = "Correlation",
     col = "#2f4569")
```

### LDA Baseline Model
```{r}
library(MASS)
library(caret)

# Make sure activity labels are factors
y_train$activity <- factor(y_train$activity)
y_test$activity <- factor(y_test$activity, levels = levels(y_train$activity))

# Start time to measure efficiency
baseline_start <- Sys.time()

# Train baseline LDA model
baseline_model <- lda(x = X_train, grouping = y_train$activity)

# Predict test labels
baseline_pred <- predict(baseline_model, X_test)$class

# End time
baseline_end <- Sys.time()

# Force predictions to have same factor levels as y_test
baseline_pred <- factor(baseline_pred, levels = levels(y_test$activity))

# Confusion matrix
baseline_cm <- confusionMatrix(data = baseline_pred,
                               reference = y_test$activity)

# baseline training accuracy (to assess theoretical upper limit of test accuracy)
baseline_train_pred <- predict(baseline_model, X_train)$class

baseline_train_pred <- factor(baseline_train_pred,
                              levels = levels(y_train$activity))

baseline_train_cm <- confusionMatrix(data = baseline_train_pred,
                                     reference = y_train$activity)

baseline_train_cm$overall["Accuracy"]

```
### Visualization of the confusion matrix
```{r}
cm <- confusionMatrix(baseline_pred, y_test$activity)
cm$table

accuracy <- cm$overall["Accuracy"]
print(accuracy)

metrics <- cm$byClass[, c("Precision", "Recall", "F1")]
metrics
```
```{r}

cm_df <- as.data.frame(cm$table)
colnames(cm_df) <- c("Predicted", "Actual", "Freq")

library(ggplot2)

ggplot(cm_df, aes(x = Actual, y = Predicted, fill = Freq)) +
  geom_tile() +
  geom_text(aes(label = Freq), color = "black", size = 4) +
  scale_fill_gradient(low = "#deebf7", high = "#08519c") +
  theme_minimal(base_size = 14) +
  labs(
    title = "Confusion Matrix (Baseline Model)",
    x = "Actual",
    y = "Predicted"
  ) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

### Principal Component Analysis

There are multiple ways to perform PCA in R; which one is the fastest?

Calculate the sample covariance matrix of `X_train`, then find the 
eigenvalues and corresponding eigenvectors and multiply `X_train` by the latter.

```{r}
scov.start <- Sys.time()
X_train.scov <- t(X_train) %*% as.matrix(X_train) / (nrow(X_train)-1)
X_train.scov.eigen <- X_train.scov %>% eigen
X_train.scov.pcs <- as.matrix(X_train) %*% X_train.scov.eigen$vectors
scov.end <- Sys.time()
X_train.scov.eivals <- X_train.scov.eigen$values
```

```{r}
(scov.end - scov.start) %>% print
```

Calculate the singular value decomposition $UDV^T$ of `X_train`, and derive the principal components $YV=UD$. 

```{r}
svd.start <- Sys.time()
X_train.svd <- X_train %>% svd
X_train.svd.pcs <- X_train.svd$u %*% diag(X_train.svd$d)
X_train.svd.eivals <- X_train.svd$d^2 / (nrow(X_train)-1)
svd.end <- Sys.time()
```

```{r}
(svd.end - svd.start) %>% print
```

Use the `PCA()` function from `FactoMineR` (We will proceed using this,
even though it is slower, because the results are more conducive to
statistical analysis)
```{r}
pca.start <- Sys.time()
X_train.pca <- PCA(X_train, ncp = nrow(X_train), graph = FALSE)
pca.end <- Sys.time()
X_train.pcs <- X_train.pca[["ind"]][["coord"]] %>% as.data.frame
colnames(X_train.pcs) <- generic_colnames
X_train.eivals <- X_train.pca$eig[,1]

X_train.v <- X_train.pca$svd$V %>% as.data.frame 
X_test.pcs <- (as.matrix(X_test) %*% as.matrix(X_train.v)) %>% as.data.frame 
colnames(X_test.pcs) <- generic_colnames 
```

```{r}
(pca.end - pca.start) %>% print
```

```{r}
X_train.pcs[1:5,1:3] %>% print
X_train.eivals[1:5] %>% print
```

Create a scatter plot of the first two principal components of the data...

```{r}
scatter2 <- plot_ly(x=X_train.pcs[,1], y=X_train.pcs[,2],
        type="scatter", mode="markers", color=y_train$activity) %>%
  layout(title = 'First Two Principal Components')

scatter2
```

and of the first three principal components.

```{r}
scatter3 <- plot_ly(x=X_train.pcs[[1]], y=X_train.pcs[[2]],
                    z=X_train.pcs[[3]],
        type="scatter3d", mode="markers", color=y_train$activity)  %>%
  layout(title = 'First Three Principal Components')

scatter3
```

How many principal components should we use in building models? Generate
a scree plot of cumulative variance.

```{r}

X_train.cum_var <- X_train.pca$eig[,3]

scree <- ggplot() +
  geom_line(aes(x=1:length(X_train.cum_var), y=X_train.cum_var)) +
  geom_point(aes(x=1:length(X_train.cum_var), y=X_train.cum_var)) +
  labs(title="Cumulative Variance Scree Plot", x="Principal Components", 
       y="% Variance Explained")

scree %>% plot
```

How many principal components do we need reach certain thresholds of variance?
```{r}
# 80%
which(X_train.cum_var < 80) %>% length %>% +1 %>% print

# 90%
which(X_train.cum_var < 90) %>% length %>% +1 %>% print

# 95%
which(X_train.cum_var < 95) %>% length %>% +1 %>% print

# 99%
which(X_train.cum_var < 99) %>% length %>% +1 %>% print

# 99.9%
which(X_train.cum_var < 99.9) %>% length %>% +1 %>% print
```

What threshold of variance do certain numbers of principal compoments
reach?

```{r}
# 1
X_train.cum_var[1] %>% print

# 5
X_train.cum_var[5] %>% print

# 10
X_train.cum_var[10] %>% print

# 20
X_train.cum_var[20] %>% print

# 50
X_train.cum_var[50] %>% print

# 100
X_train.cum_var[100] %>% print

# 200
X_train.cum_var[200] %>% print
```
Which variables contribute most to the first principal component?
```{r}
X_train.pc1_cont <- X_train.v[,1] %>% sort(index.return=TRUE, decreasing=TRUE)

t5_ind <- 1:5

X_train.pc1_cont_t5 <- X_train.pc1_cont$ix[t5_ind]

X_train.pc1_cont_t5 %>% print
feature_descs$desc[X_train.pc1_cont_t5] %>% print
X_train.v[X_train.pc1_cont_t5,1] %>% print
```

Which variables contribute least to the first principal component?
```{r}
b5_ind <- (features-4):features
X_train.pc1_cont_b5 <- X_train.pc1_cont$ix[b5_ind]

X_train.pc1_cont_b5 %>% print
feature_descs$desc[X_train.pc1_cont_b5] %>% print
X_train.pc1_cont$x[b5_ind] %>% print
```

Plot the coefficients of each variable in the first principle component
in decreasing order.

```{r}
pc1_plot <- ggplot() +
  geom_line(aes(x=1:length(X_train.pc1_cont$x), y=X_train.pc1_cont$x)) +
  geom_point(aes(x=1:length(X_train.pc1_cont$x), y=X_train.pc1_cont$x)) +
  labs(title="First Principal Component Coefficients", x="Rank", 
       y="Coefficient")

pc1_plot %>% plot
```

### PCA components for transformation

We chose the first PCs explaining at least 95% of variance of the original features.

```{r}
pcs_95 <- which(X_train.cum_var >= 95)[1]
pcs_95
```

Using these, subset the training and test sets using these first PCs (note: no leakage, only training set used to compute loadings).

```{r}
X_train.pcs_95 <- X_train.pcs[, 1:pcs_95]
X_test.pcs_95 <- X_test.pcs[, 1:pcs_95]
```

### LDA on PCA transformed features

After transforming the features using PCA, we fit another LDA model.

We compare all metrics including precision, recall, f1, accuracy, and runtime against the original model.

Note To Professor: The "nicer tables" included in the presentation will only load if you use library("gt"). The package will not install automatically when knitting, so the code will be left but commented out. The knitr tables should produce the same results as our presentation (minus runtime differences by hardware).

```{r}
# load needed libraries
library(MASS)
library(caret)

# refactoring due to errors encountered
activity_levels <- c("LAYING", "SITTING", "STANDING", "WALKING",
                     "WALKING_DOWNSTAIRS", "WALKING_UPSTAIRS")
y_train$activity <- factor(as.character(y_train$activity),
                           levels = activity_levels)
y_test$activity <- factor(as.character(y_test$activity),
                           levels = activity_levels)

# computational efficiency start
eff_start <- Sys.time()

# training the model on the transformed dataset on 102 components
pca_lda <- lda(x = X_train.pcs_95, grouping = y_train$activity)

# create predictions
pca_lda_pred <- predict(pca_lda, X_test.pcs_95)$class

# computational efficiency end
eff_end <- Sys.time()

# create predictions on transformed test set
pca_lda_pred <- factor(as.character(pca_lda_pred),
                       levels = activity_levels)

# confusion matrix
cm_pca_lda <- confusionMatrix(pca_lda_pred,
                              reference = y_test$activity)
cm_pca_lda

# convert into a dataframe for viewing
cm_pca_lda_df <- as.data.frame(cm_pca_lda$table)
colnames(cm_pca_lda_df) <- c("Predicted", "Actual", "Freq")

# visualize confusion matrix, setup from earlier copied exactly
ggplot(cm_pca_lda_df, aes(x = Actual, y = Predicted, fill = Freq)) +
  geom_tile() +
  geom_text(aes(label = Freq), color = "black", size = 4) +
  scale_fill_gradient(low = "#deebf7", high = "#08519c") +
  theme_minimal(base_size = 14) +
  labs(
    title = "Confusion Matrix (PCA Transformed Model)",
    x = "Actual",
    y = "Predicted"
  ) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

# percent of original features
print(pcs_95 / ncol(X_train_raw) * 100)

# Find the metrics
print(cm_pca_lda$overall["Accuracy"])
print(cm_pca_lda$byClass[, c("Precision", "Recall", "F1")])

# save for viewing
accuracy2 <- cm_pca_lda$overall["Accuracy"]
metrics2 <- cm_pca_lda$byClass[, c("Precision", "Recall", "F1")]

# After PCA time
time2 <- eff_end - eff_start
print(time2)

# before PCA time
time1 <- baseline_end - baseline_start
print(time1)

# convert into a usable format for calculation
time1 <- as.numeric(time1, unit = "secs")
time2 <- as.numeric(time2, unit = "secs")

# relative speedup
print((time1-time2) / time1 * 100)


# print accuracy before and after
print(accuracy)
print(accuracy2)

# assemble before and after in a dataframe
baseline_metrics <- data.frame(
  Precision = round(metrics[, "Precision"], 3),
  Recall = round(metrics[, "Recall"], 3),
  F1 = round(metrics[, "F1"], 3),
  Runtime = round(time1, 3)
)

after_metrics <- data.frame(
  Precision = round(metrics2[, "Precision"], 3),
  Recall = round(metrics2[, "Recall"], 3),
  F1 = round(metrics2[, "F1"], 3),
  Runtime = round(time2, 3)
)

# knitr
library(knitr)

# gt
# library("gt")

# print with knitr
knitr::kable(baseline_metrics, caption = "LDA metrics before PCA")

# nicer table
baseline_metrics$Class <- rownames(baseline_metrics)
baseline_metrics$Class <- gsub("Class: ", "", baseline_metrics$Class)
baseline_metrics <- baseline_metrics[, c("Class", "Precision", "Recall", "F1", "Runtime")]

#baseline_metrics %>%
#  gt()

knitr::kable(after_metrics, caption = "LDA metrics after PCA")

# nicer table
after_metrics$Class <- rownames(after_metrics)
after_metrics$Class <- gsub("Class: ", "", after_metrics$Class)
after_metrics <- after_metrics[, c("Class", "Precision", "Recall", "F1", "Runtime")]

#after_metrics %>%
#  gt()

print(baseline_metrics)

# F1 comparison
round(metrics2[, "F1"] - metrics[, "F1"], 3)
print(summary(round(metrics2[, "F1"] - metrics[, "F1"], 3)))

# Model comparison dataframe
model_comparison <- data.frame(
  Metric = c("Features", "Accuracy", "Mean F1", "Runtime (sec)"),
  Baseline = c(as.character(ncol(X_train_raw)),
       round(as.numeric(accuracy), 3),
       round(mean(metrics[, "F1"]), 3),
       round(time1, 3)),
  PCA = c(as.character(ncol(X_train.pcs_95)),
       round(as.numeric(accuracy2), 3),
       round(mean(metrics2[, "F1"]), 3),
       round(time2, 3)),
  Change = c(
       paste0(round((ncol(X_train.pcs_95) - ncol(X_train_raw)) / ncol(X_train_raw) * 100, 3), "%"),
       round(as.numeric(accuracy2) - as.numeric(accuracy), 3),
       round(mean(metrics2[, "F1"]) - mean(metrics[, "F1"]), 3),
       paste0(round((time2 - time1) / time1 * 100, 3), "%")
  )
)

# print with knitr
knitr::kable(model_comparison, caption = "Metric Comparison Table") 

# nicer table
#model_comparison %>%
#  gt()

```

### k-Nearest Neighbors



What combination of principal components and $k$ yields the best test
accuracy? 

```{r}

knn_test_acc <- function(k, pcs) { 
  test_preds <- knn(train=X_train.pcs[1:pcs], test=X_test.pcs[1:pcs],
                    cl=y_train$index, k = k)
  return (length(which(y_test$index == test_preds)) / test_len) 
} 

min_pcs <- 1
max_pcs <- 301
pc_interval = 30

min_k <- 1
max_k <- 401
k_interval = 40

coarse_knn_grid <- expand.grid(pcs=seq(min_pcs, max_pcs, pc_interval),
                               k=seq(min_k, max_k, k_interval))

for (row in rownames(coarse_knn_grid)) {
  pcs <- coarse_knn_grid[row,]$pcs
  k <- coarse_knn_grid[row,]$k
  coarse_knn_grid[row, "test_acc"] <- knn_test_acc(k, pcs)
}

coarse_knn <- ggplot(coarse_knn_grid, aes(x = k, y = pcs, fill = test_acc)) +
  geom_tile()  +
  labs(title="Coarse Hyperparameter Grid Search (KNN)", x="k", 
       y="Principal Components", fill="Test Accuracy") +
  coord_cartesian(xlim = c(min_k, tail(coarse_knn_grid, 1)$k), 
                  ylim = c(min_pcs, tail(coarse_knn_grid, 1)$pcs))

coarse_knn %>% plot
```

```{r}
coarse_knn_grid[which.max(coarse_knn_grid$test_acc),] %>% print
```

$k$ and the number of principal components seem to be independent in determining the test accuracy of the model. Thus, we can save time by simply tuning $k$ at some relatively low number of principal components.

```{r}
fine_k <- 1:100

fine_k_acc <- sapply(fine_k, knn_test_acc, pcs=10)

fine_k_knn <- ggplot() +
  geom_line(aes(x=fine_k, fine_k_acc)) +
  labs(title="Fine-Tuning k (10 PCs)", x="k", 
       y="Test Accuracy")

fine_k_knn %>% plot
```

```{r}
which.max(fine_k_acc) %>% print
fine_k_acc[which.max(fine_k_acc)] %>% print
```

Plot the test accuracy at different numbers of principal components for
$k=20$.

```{r}
fine_pcs <- 1:300

fine_pcs_acc <- sapply(fine_pcs, knn_test_acc, k=20)

fine_pcs_knn <- ggplot() +
  geom_line(aes(x=fine_pcs, fine_pcs_acc)) +
  labs(title="Fine-Tuning Principal Components (k=20)", x="Principal Components", 
       y="Test Accuracy")

fine_pcs_knn %>% plot
```

```{r}
which.max(fine_pcs_acc) %>% print
fine_pcs_acc[which.max(fine_pcs_acc)] %>% print
```

How does this compare to performing KNN on the original dataset?

```{r}
full_knn.start <- Sys.time()
full_knn_preds <- knn(train=X_train, test=X_test,
                   cl=y_train$index, k = 20)
full_knn.end <- Sys.time()
length(which(y_test$index == full_knn_preds)) / test_len %>% print
(full_knn.end - full_knn.start) %>% print
```

```{r}
pca_knn.start <- Sys.time()
pca_knn_preds <- knn(train=X_train.pcs[1:120], test=X_test.pcs[1:120],
                   cl=y_train$activity, k = 20)
pca_knn.end <- Sys.time()
(pca_knn.end - pca_knn.start) %>% print
```


Create a confusion matrix and generate descriptive statistics.


```{r}
knn_cm <- confusionMatrix(pca_knn_preds, as.factor(y_test$activity))

knn_cm_df <- as.data.frame(knn_cm$table)
colnames(knn_cm_df) <- c("Predicted", "Actual", "Freq")

knn_cm_plot <- ggplot(knn_cm_df, aes(x = Actual, y = Predicted, fill = Freq)) +
  geom_tile() +
  geom_text(aes(label = Freq), color = "black", size = 4) +
  scale_fill_gradient(low = "#deebf7", high = "#08519c") +
  theme_minimal(base_size = 14) +
  labs(
    title = "Confusion Matrix (KNN)",
    x = "Actual",
    y = "Predicted"
  ) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

knn_cm_plot %>% plot
```

```{r}
knn_cm$overall[c("Accuracy")] %>% print

knn_metrics <- knn_cm$byClass[, c("Precision", 
                                  "Recall", "F1")]
knn_metrics %>% print
```